---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

## nlsLoop

Tools for expanding the non-linear regression method nls and nlsList from nlme.

### Issues and suggestions

Please report any issues/suggestions in the [issues link](https://github.com/padpadpadpad/nlsLoop/issues) for the repository.

Travis: [![Build Status](https://travis-ci.org/padpadpadpad/nlsLoop.svg?branch=master)](https://travis-ci.org/padpadpadpad/nlsLoop)

### Overview

The mainstay of this package is `nlsLoop::nlsLoop()`. If you have a dataset where you want to fit the same model over many levels of a factor using non-linear regression, you may use `nlme::nlsList()`. However, `nlsList()` only allows for one set of starting values so it is possible that not all of the models will converge if there is large variance in the data or the shape of the relationships are different, and thus the parameter values, are likely to be very different.

`nlsLoop()` allows for a range of starting values, looping through and attempting a fit for thousands of unique combination of starting values at each level of the factor, picking the best fit for each model using AIC scores.

### Tutorial

A more in-depth tutorial and explanation of how to use `nlsLoop()` can be found by calling `vignette('nlsLoop')` from within R.

### Installation and examples

#### 1. Installation

```{r install, message=FALSE, eval = FALSE}
# install package
devtools::install_github("padpadpadpad/nlsLoop", build_vignettes = TRUE)

```

#### 2. Run nlsLoop()

```{r run_nlsLoop, message=FALSE, warning=FALSE, results='hide'}

# load in nlsLoop and other packages
library(nlsLoop)
library(ggplot2)

# load nlsLoop vignette
vignette('nlsLoop')

# load in example data set
data("Chlorella_TRC")

# define the Sharpe-Schoolfield equation
schoolfield_high <- function(lnc, E, Eh, Th, temp, Tc) {
  Tc <- 273.15 + Tc
  k <- 8.62e-5
  boltzmann.term <- lnc + log(exp(E/k*(1/Tc - 1/temp)))
  inactivation.term <- log(1/(1 + exp(Eh/k*(1/Th - 1/temp))))
  return(boltzmann.term + inactivation.term)
  }

# run nlsLoop
fits <- nlsLoop(ln.rate ~ schoolfield_high(lnc, E, Eh, Th, temp = K, Tc = 20),
                     data = Chlorella_TRC,
                     tries = 500,
                     id_col = 'curve_id',
                     param_bds = c(-10, 10, 0.1, 2, 0.5, 5, 285, 330),
                     r2 = 'Y',
                     supp_errors = 'Y',
                     AICc = 'Y',
                     na.action = na.omit,
                     lower = c(lnc = -10, E = 0, Eh = 0, Th = 0))
```

#### 3. Check output

```{r look_at_data}
head(fits$params)

head(fits$predictions)
```

#### 4. Check fit of single curve

```{r first_fit_plot, fig.height=6, fig.width=7}
# plot a single curve
plot_id_nlsLoop(data = Chlorella_TRC, param_data = fits, id = '1')
```

#### 5. Check fit of all curves (creates a pdf)

```{r pdf_fits, eval=FALSE}
# create pdf of each curve
plot_all_nlsLoop('path/of/where/you/want/to/save/me.pdf', raw_data = Chlorella_TRC, param_data = fits)
```

#### 6. Plot predictions

```{r data_wrangling, fig.height=7,  fig.width = 7}
# get distinct values of process, flux and growth.temp for each value of curve_id
d_treatment <- Chlorella_TRC[,c('curve_id','process', 'growth.temp', 'flux')]
d_treatment <- d_treatment[!duplicated(d_treatment),]

# merge with predictions by curve_id
fits$predictions <- merge(fits$predictions, d_treatment, by = 'curve_id')

# plot every curve
ggplot() +
  geom_point(aes(K, ln.rate, col = flux), size = 2, Chlorella_TRC) +
  geom_line(aes(K, ln.rate, col = flux, group = curve_id), alpha = 0.5, fits$predictions) +
  facet_wrap(~ growth.temp + process, labeller = labeller(.multi_line = F)) +
  scale_colour_manual(values = c('green4', 'black')) +
  theme_bw(base_size = 12, base_family = 'Helvetica') +
  ylab('log Metabolic rate') +
  xlab('Assay temperature (ºC)') +
  theme(legend.position = c(0.9, 0.15))

```


#### 7. Get and plot confidence intervals of parameters

```{r, confint_nlsLoop, fig.width = 7, fig.height = 8}
# calculate confidence intervals for each fit
CIs <- confint_nlsLoop(Chlorella_TRC, fits)

# bind with factors dataframe
CIs <- merge(CIs, d_treatment, by = 'curve_id')

# plot
ggplot(CIs, aes(col = flux)) +
  geom_point(aes(curve_id, mean)) +
  facet_wrap(~ param, scale = 'free_x', ncol = 4) +
  geom_linerange(aes(curve_id, ymin = CI_lwr, ymax = CI_upr)) +
  coord_flip() +
  scale_color_manual(values = c('green4', 'black')) +
  theme_bw(base_size = 12, base_family = 'Helvetica') +
  theme(legend.position = 'top') +
  xlab('curve') +
  ylab('parameter estimate')



```
